---
header-includes: |
  \usepackage{float}
  \usepackage{wrapfig}
  \usepackage{fancyhdr}
  \usepackage{lastpage}
  \usepackage{afterpage}
  \usepackage{fvextra}
  \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
output:
  pdf_document:
    highlight: tango
    dev: png
    fig_caption: yes
    number_sections: yes
    extra_dependencies: ["eso-pic","graphicx","transparent", "xcolor", "floatflt", "listings", "lipsum"]
    latex_engine: xelatex ## adds support for non unicode characters
      
---
\definecolor{darkblue}{RGB}{38,50,90}

<!-- Full document text watermark -->
<!-- \AddToShipoutPictureFG{ -->
<!--   \AtPageCenter{% or \AtTextCenter -->
<!--     \makebox[0pt]{\rotatebox[origin=c]{45}{% -->
<!--     \scalebox{10}{\texttransparent{0.2}{DRAFT DRAFT DRAFT DRAFT DRAFT DRAFT DRAFT DRAFT DRAFT DRAFT }}% -->
<!--     }} -->
<!--   } -->
<!-- } -->


<!-- Full document image watermark -->
<!-- \AddToShipoutPictureFG{ -->
<!--         \put(-200,-25){ -->
<!--         \parbox[b][\paperheight]{\paperwidth}{% -->
<!--             \vfill -->
<!--             \centering{ -->
<!--             \transparent{0.05}\includegraphics[angle=45,width=2\textwidth]{images/pepe.PNG} -->
<!--             }% -->
<!--             \vfill -->
<!--         } -->
<!--     } -->
<!-- } -->




<!-- dump the code for the apendix -->


<!-- Prepare title info -->
```{r titles, include=FALSE, echo=FALSE}
Title <- "732A97 Multivariate Statistical Methods"
Subtitle <- "Assignment 1 Examining multivariate data"
Author <- "Alexander Karlsson(aleka769),Raymond Sseguya(y),Mariano Mariani(marma330), Ruben Munoz(rubmu773)"
Date <- Sys.Date()
Chapter01 <- "Describing individual variables"
Chapter02 <- "Relationships between the variables"
Chapter03 <- "Examining for extreme values"
```
<!-- To keep a simple main page -->
\pagestyle{empty}

<!-- Start main page -->
\noindent\null\vspace{2 cm}

\noindent\hrulefill{\hrule height 0.6pt}\null\vspace{-0.2 cm}

\begin{center}

\Huge{\bf `r Title`}\vspace{.5 cm}

\LARGE{`r Subtitle`}\vspace{0 cm}

\normalsize{\textit{`r Author`}}\vspace{0 cm}

\normalsize{\textit{`r Date`}}\vspace{-0.5 cm}

\end{center}
 
\noindent\hrulefill{\hrule height 0.6pt}\null\vspace{2 cm}

\tableofcontents

\newpage
<!-- line is very touchy with spaces before and after -->

<!-- Style for the rest of the document -->
\pagestyle{fancy}
\fancyhead[LH]{\textsc{`r Title`}}
\fancyfoot[LF]{\textsl{`r Subtitle`}}

\fancyfoot[CF]{\thepage\ of \pageref{LastPage}}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\footrulewidth}{0.4pt}

<!-- \twocolumn -->

```{r child='01.Rmd'}
```

\newpage

 ```{r child='02.Rmd'}
 ```

\newpage

 ```{r child='03.Rmd'}
 ```
 
\newpage

```{r child='apendix.Rmd'}
```