---
title: "Lab 3 732A97"
author: "Raymond Sseguya"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

# Data
```{r}
library(dplyr)

trackrcs <- read.table("T1-9.dat",
  col.names = c("countries", "x100m", "x200m", 
        "x400m", "x800m", "x1500m", "x3000m", "marathon"))
rownames(trackrcs) <- (trackrcs)[,1]

```


# Question 1: Principal components, including interpretation of them

## a) Obtain the sample correlation matrix R for these data, and determine its eigenvalues and eigenvectors.
```{r}
S <- cov((trackrcs)[,-1])
R <- cov2cor(S); R

eigen(R)$values
eigen(R)$vectors

```


## b) Determine the first two principal components for the standardized variables. Prepare a table showing the correlations of the standardized variables with the components, and the cumulative percentage of the total (standardized) sample variance explained by the two components.

### the first two principal components for the standardized variables
```{r}
res=prcomp((trackrcs)[,-1], scale. = FALSE)
# No sacling at this point because we are going to use the correlation matrix later

# Each PC is a linear combination of the original variables
#### res$rotation
res$rotation[,1:2]
```


\pagebreak

### correlation of the standardized variables with the components
```{r}
### Method 1 (based on textbook)
eigenvalues=res$sdev^2
CorWithPC <-
  t(res$rotation[,1:2])%*%sqrt(diag(eigenvalues))%*%solve(sqrt(diag(diag(R))))
colnames(CorWithPC) <- colnames(trackrcs[,-1])
t(CorWithPC)

```


### cummulative percentage of total standardized sample variance explained by the 2 components
```{r}
CorWithPC %>% apply(MARGIN=1,FUN=abs) %>% t() %>%
  apply(MARGIN=1,FUN=function(a) 100*cumsum(a)/sum(a))

```



\pagebreak

## c) Interpret the two principal components obtained in Part b. (Note that the first component is essentially a normalized unit vector and might measure the athletic excellence of a given nation. The second component might measure the relative strength of a nation at various running distances.)



\pagebreak

## d) Rank the nations based on their score on the first principal component. Does this ranking correspond with your inituitive notion of athletic excellence for the various countries?
```{r}
### Method 2 (based on observation)
CorWithPC2 <- cor( t(scale((trackrcs)[,-1])),  res$rotation[,1:2] )

countries_CorWithPC = cbind.data.frame(countries = trackrcs[,1], 
                            correlation = (CorWithPC2[,1]),
                            unsigned_correlation = abs(CorWithPC2[,1]))
                              
countries_CorWithPC[,1][order(countries_CorWithPC[,3], decreasing = TRUE)]
countries_CorWithPC[,1][order(countries_CorWithPC[,2], decreasing = TRUE)]
```
We get a different ranking when we use correlations directly (with their signs) from when we use unsigned correlations (the magnitude of the correlations). The ranking seems to be very innacurate about the athletic excellence for the various countries.



\pagebreak


# Question 2: Factor analysis
```{r}
factanal(trackrcs[,-1], factors = 3, covmat = S) # varimax is the default
factanal(trackrcs[,-1], factors = 3, covmat = R)

psych::principal(cov2cor(S), nfactors=3, rotate="varimax")
psych::fa(cov2cor(S), nfactors=3, rotate="varimax")

```

